---
title: "PA1_template.Rmd"
author: "Kumar Saurabh"
date: "June 22, 2017"
output: html_document
---

##Code for reading in the dataset and/or processing the data
```{r}
# graphics library load
library(ggplot2)
# Read base data into a data frame.
activitydata<-read.csv("activity.csv")
# change date type to date class 
activitydata$date <- as.Date(as.character(activitydata$date))
# create NA logical vector
activitydataNA <- is.na(activitydata$steps)
# create clean base for later mean calculations
cleanactivitydata <- activitydata[!activitydataNA,]
# aggregate clean non NA steps per day (SUM)
activitydataByDay <- aggregate(activitydata$steps, by=list(activitydata$date), sum)
# adjust column names
names(activitydataByDay)[1] ="date"
names(activitydataByDay)[2] ="totalsteps"
```

## Histogram of the total number of steps taken each day

```{r}
# Plot using ggplot
ggplot(activitydataByDay, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")
```

## Mean and median number of steps taken each day
```{r}
#Mean of steps taken per day
mean(activitydataByDay$totalsteps,na.rm=TRUE)
#median of steps taken per day
median(activitydataByDay$totalsteps,na.rm=TRUE)
```



##Time series plot of the average number of steps taken
```{r}
# Get mean data by interval
MeanDataByInterval <- aggregate(cleanactivitydata$steps, by=list(cleanactivitydata$interval), mean)
# set the column names
names(MeanDataByInterval)[1] ="interval"
names(MeanDataByInterval)[2] ="steps"

ggplot(MeanDataByInterval, aes(x = interval, y=steps)) +
  labs(title = "Sum of Steps by Interval", x = "interval", y = "steps")+
  geom_line(color="red") 
```

##The 5-minute interval that, on average, contains the maximum number of steps
```{r}
maxInterval <- MeanDataByInterval[which.max(MeanDataByInterval$steps),]
maxInterval
```

##Code to describe and show a strategy for imputing missing data
```{r}
# generate listing of NA's
missingVals <- sum(activitydataNA)
missingVals
# new dataset
baseData2 <- activitydata
# NA's in dataset
# dataset minus NA's for Mean calculation
NABase2 <- baseData2[is.na(baseData2$steps),]
cleanBase2 <- baseData2[!is.na(baseData2$steps),]
# generate Mean Data2 by interval
MeanData2ByInterval <- aggregate(cleanBase2$steps, by=list(cleanBase2$interval), sum)
names(MeanData2ByInterval)[1] ="interval"
names(MeanData2ByInterval)[2] ="steps"
#Imput Method
baseData2 <- activitydata
missingData <- is.na(baseData2$steps)
meanVals <- tapply(cleanactivitydata$steps, cleanactivitydata$interval, mean, na.rm=TRUE, simplify=TRUE)

baseData2$steps[missingData]<-meanVals[as.character(  baseData2$interval[missingData])]
                                        
# original missing data count
sum(missingData)
# count of NA values
sum(is.na(baseData2$steps))
```


##Histogram of the total number of steps taken each day after missing values are imputed
```{r}
FullSummedDataByDay <- aggregate(baseData2$steps, by=list(baseData2$date), sum)

names(FullSummedDataByDay)[1] ="date"
names(FullSummedDataByDay)[2] ="totalsteps"
# Plot using ggplot
ggplot(FullSummedDataByDay, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")
# Mean on New Data
mean(FullSummedDataByDay$totalsteps)
# Median on New Data
median(FullSummedDataByDay$totalsteps)
#Yes, he mean is the same but the median has risen 1.19 steps.
#The effect of using mean data per interval as a data impute method for missing values seems to push overall data towards the mean.
```

##Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r}
baseData2$weekday <- weekdays(baseData2$date)
baseData2$weekend <- ifelse (baseData2$weekday == "Saturday" | baseData2$weekday == "Sunday", "Weekend", "Weekday")
#baseData2$weekend <- as.factor(baseData2$weekend)
head(baseData2,5)
MeanDataWeekendWeekday <- aggregate(baseData2$steps, by=list(baseData2$weekend, baseData2$interval), mean)
names(MeanDataWeekendWeekday)[1] ="weekend"
names(MeanDataWeekendWeekday)[2] ="interval"
names(MeanDataWeekendWeekday)[3] ="steps"

ggplot(MeanDataWeekendWeekday, aes(x = interval, y=steps, color=weekend)) +
  geom_line() +
  facet_grid(weekend ~ .) +
  labs(title = "Mean of Steps by Interval", x = "interval", y = "steps")
#There seems to be variation in the beginning of the day during weekdays, likely due to workplace activities. There seems to be an overall slightly larger incidence of steps during the weekends
```


